sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-11-23 doc/paper.Rmd
Add timing fig

```{r diagram, fig.cap="(ref:diagram-fig)"}
if (file.exists(here::here("figs/diagram.pdf"))) {
  knitr::include_graphics(here::here("figs/diagram.pdf"))
} else {
  plot(1, type = "n", main = "figs/diagram.pdf missing; will be synced when complete")
}
```

@@

```{r}
g_pred <- ggplot(bc_coast_proj) + geom_sf(
                                           # colour=NA, fill="white"
                                           fill=NA, colour="gray30"
                                           # fill="gray70", colour=NA
                                          ) +
  geom_raster(data = p, aes(x = X * 1000, y = Y * 1000, fill = exp(est))) +
  coord_sf(datum = st_crs(utm_zone9)) +
  scale_x_continuous(limits = c(230957.7 + 105000, 1157991 - 570000),
    labels = function(x) paste0(x/1000)
  ) +
  scale_y_continuous(limits = c(5366427 + 270000, 6353456 - 513000),
    labels = function(x) paste0(x/1000)
  ) +  
  scale_fill_viridis_c(trans = "sqrt", option = "C") +
  # scale_fill_viridis_c(trans = "sqrt", option = "B", begin = 0.3) +
  labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
  theme(
    # panel.background = element_rect(fill = "slategray1", colour= "white"), 
    legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt"))
```

@@

```{r pcod-fig, fig.cap="(ref:pcod-fig-cap)", fig.width=9, out.width="5.2in", fig.asp=1, fig.align='center'}
g_omega <- ggplot(bc_coast_proj) + geom_sf(
                                           # colour=NA, fill="white"
                                           fill=NA, colour="gray30"
                                           # fill="gray70", colour=NA
                                           ) +
  geom_raster(data = p, aes(x = X * 1000, y = Y * 1000, fill = omega_s)) +
  coord_sf(datum = st_crs(utm_zone9)) +
  scale_x_continuous(limits = c(230957.7 + 105000, 1157991 - 570000),
    labels = function(x) paste0(x/1000)
  ) +
  scale_y_continuous(limits = c(5366427 + 270000, 6353456 - 513000),
    labels = function(x) paste0(x/1000)) +
  # scale_fill_gradient2(high = reds[4], low = scales::muted("purple")) +
  scale_fill_gradient2(high = reds[4], low = blues[4]) +
  labs(fill = "Spatial random\nfield deviations", x = "UTMs E (km)", y = "UTMs N (km)") +
  theme(
    # panel.background = element_rect(fill = "slategray1", colour= "white"),
    legend.position = c(0.2, 0.2), legend.key.height = unit(10, 'pt'))

cowplot::plot_grid(g_mesh, g_smooth, g_omega, g_pred, labels = "auto")
```

@@

@@

inlabru is similar to sdmTMB in design, but differs fundamentally in that fitting is achieved via INLA and approximate Bayesian computation.
For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru [@osgood-zimmerman2021], but in our experience, those differences depend on many aspects of the data, SPDE mesh, model, and fitting algorithm settings (TODO Supp fig?).
One major capability included in sdmTMB but not INLA/inlabru is anisotropy.
Furthermore, we think the sdmTMB interface will be more familiar than inlabru to users of `glm()`, lme4, mgcv, or glmmTMB.
On the other hand, inlabru fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with INLA.

mgcv can fit many similar models to sdmTMB either with various flavours of splines or even with the SPDE approximation to random fields (REF); however, there are some notable advantages of sdmTMB for the specific class of models discussed here.
First, the smoothers typically used with mgcv (thin plate splines, ...), while capable of making equivalent predictions (REF) do not offer the ecological interpretability of a spatially explicit random field model (spatial range and variance).
Second, the implementation of the Tweedie distribution in mgcv requires the power parameter to be fixed in a mixed effects model.
Third, mgcv does not have built-in functionality to calculate derived values with uncertainty such as population indexes or center of gravity.
Despite these sdmTMB advantages for the models discussed here, mgcv is an immensely powerful modelling package that extends well beyond spatial random field models.

>>>>

```{r diagram, fig.cap="(ref:diagram-fig)"}
f <- here::here("figs", "diagram.pdf")
if (!file.exists(f)) {
  download.file("https://www.dropbox.com/s/j8km3hv8xz9b63k/diagram.pdf?dl=1", 
    destfile = f)
}
knitr::include_graphics(f)
```

@@

```{r}
g_pred <- ggplot(bc_coast_proj) +
  geom_sf(fill = NA, colour = "gray30", lwd = 0.3) +
  geom_raster(data = p, aes(x = X * 1000, y = Y * 1000, fill = exp(est))) +
  coord_sf(datum = st_crs(utm_zone9)) +
  scale_x_continuous(
    limits = c(230957.7 + 105000, 1157991 - 570000),
    labels = function(x) paste0(x / 1000)
  ) +
  scale_y_continuous(
    limits = c(5366427 + 270000, 6353456 - 513000),
    labels = function(x) paste0(x / 1000)
  ) +
  scale_fill_viridis_c(trans = "sqrt", option = "C") +
  # scale_fill_viridis_c(trans = "sqrt", option = "B", begin = 0.3) +
  labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
  theme(
    # panel.background = element_rect(fill = "slategray1", colour= "white"),
    legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt")
  )
```

@@

```{r pcod-fig, fig.cap="(ref:pcod-fig-cap)", fig.width=9, out.width="5.2in", fig.asp=1, fig.align='center'}
g_omega <- ggplot(bc_coast_proj) +
  geom_sf(fill = NA, colour = "gray30", lwd = 0.3) +
  geom_raster(data = p, aes(x = X * 1000, y = Y * 1000, fill = omega_s)) +
  coord_sf(datum = st_crs(utm_zone9)) +
  scale_x_continuous(
    limits = c(230957.7 + 105000, 1157991 - 570000),
    labels = function(x) paste0(x / 1000)
  ) +
  scale_y_continuous(
    limits = c(5366427 + 270000, 6353456 - 513000),
    labels = function(x) paste0(x / 1000)
  ) +
  # scale_fill_gradient2(high = reds[4], low = scales::muted("purple")) +
  scale_fill_gradient2(high = reds[4], low = blues[4]) +
  labs(fill = "Spatial random\nfield deviations", x = "UTMs E (km)", y = "UTMs N (km)") +
  theme(
    # panel.background = element_rect(fill = "slategray1", colour= "white"),
    legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt")
  )
cowplot::plot_grid(g_mesh, g_smooth, g_omega, g_pred, labels = "auto")
```

@@

## Speed comparison

We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting a SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \@ref(fig:timing)).
In this test, for mesh resolutions under 500 nodes, sdmTMB fit the models the fastest.
inlabru was the least affected by mesh resolution and beyond 500 nodes was faster than sdmTMB with default sdmTMB settings.
mgcv was slightly slower than sdmTMB and was more affected by mesh resolution.
With the random effect normalizing constant calculated in R (`normalize = TRUE`), sdmTMB was the fastest across all mesh resolutions (note that this option may not be possible or as fast for all models).
Results with up to an order of magnitude more data points slightly shift the lines up without affecting the overall pattern (not shown).

(ref:timing-fig) Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
Lines represent means and ribbons ranges across 20 random iterations.
Note the log x and y axes.
sdmTMB used the default settings and `normalize = TRUE`, which calculates the random effect normalizing constant in R.
inlabru used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
mgcv used `bam()`, `method = fREML`, and discretized covariates [@miller2019].
inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library.
All platforms were restricted to one core and could be faster with parallel computation.
<!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->

```{r timing, fig.cap="(ref:timing-fig)", out.width="3.5in", fig.align='center'}
f <- here::here("figs", "timing2-logx.pdf")
if (!file.exists(f)) {
  download.file("https://www.dropbox.com/s/3aco85mizuplz7z/timing2-logx.pdf?dl=1", 
    destfile = f)
}
knitr::include_graphics(f)
```

@@

inlabru is similar to sdmTMB in design, but differs fundamentally in that fitting is achieved via INLA and approximate Bayesian computation.
For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru, but those differences depend on the SPDE mesh, model, algebra libraries, and fitting algorithm settings [Fig. \@ref(fig:timing), @osgood-zimmerman2021].
One major capability included in sdmTMB but not INLA/inlabru is anisotropy.
Furthermore, we think the sdmTMB interface will be more familiar than inlabru to users of `glm()`, lme4, mgcv, or glmmTMB.
On the other hand, inlabru fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with INLA.

mgcv can fit many similar models to sdmTMB either with various flavours of splines or even with the SPDE approximation to random fields [@miller2019]; however, there are some notable advantages of sdmTMB for the specific class of models discussed here.
First, the smoothers typically used with mgcv (thin plate splines, ...), while capable of making equivalent predictions [@miller2019; @pedersen2019] do not offer the ecological interpretability of a spatially explicit random field model (spatial range and variance).
Second, the implementation of the Tweedie distribution in mgcv requires the power parameter to be fixed in a mixed effects model.
Third, mgcv does not have built-in functionality to calculate derived values with uncertainty such as population indexes or center of gravity.
Despite these sdmTMB advantages for the models discussed here, mgcv is an immensely powerful modelling package that extends well beyond spatial random field models.
